Customizing Software by Direct Manipulation of Structured Data

### 2020-05-20 paper.tex
edit intro

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Table-Driven Customization of Web Applications}

@@

For decades, we have known about an easier model: \emph{direct
manipulation} \citep{shneiderman1983}, where ``visibility of the object
of interest'' replaces ``complex command language syntax''. Direct
manipulation is widely used in GUIs today, but rarely applied for
software customization. In this work, we ask: what would it look like to
build a software customization interface that deeply relies on direct
manipulation?

We take particular inspiration from spreadsheets and visual database
query interfaces \citep{2020a, bakke2016}. These tools enable users to
construct queries, edit data, and compute derived values through direct
manipulation of the data itself, with intermediate feedback available at
every step. These tools have successfully enabled millions of end users
to view, edit, and compute with data.

We have developed a technique called \emph{table-driven customization}
which applies ideas from visual query interfaces in the context of
software customization. We provide the end user with a table view where
they can directly manipulate the internal state of an application.
Changes in the structured table view result in changes to the original
user interface of the application, enabling the user to customize an
application with immediate feedback.

To demonstrate the viability of this approach, we have developed a
browser extension called Wildcard which implements table-driven
customization in the context of the Web. Wildcard works with existing
web applications by using web scraping techniques to extract relevant
application state into a table and to reify changes back into the
original application UI. In Section~\ref{sec:examples} we present
examples of using Wildcard to implement useful customizations across
many real websites, ranging from sorting lists of data to adding whole
new features to applications.

In Section~\ref{sec:architecture} we explain the architecture of the
Wildcard system, focusing on the key abstraction of the \emph{table
adapter} which manages a single table of data. We explain why some of
these abstractions are general enough to support many customizations
beyond the ones we've built so far in Wildcard---for example, we explore
how applications themselves could be re-architected to expose tabular
data to end users, rather than relying on web scraping techniques.

@@

\begin{itemize}
\tightlist
\item
  \emph{Data-oriented programming}: End users should be able to
  customize an application by examining and modifying its data, rather
  than by issuing imperative commands to an API
\item
  \emph{Black-box + semantic}: Typically, ``black-box'' customization
  tools, which don't rely on official extension APIs, resort to offering
  low-level APIs for customization. Instead, we suggest a third-party
  community of semantic wrappers, offering end users a more convenient
  semantic API for customization, even if it wasn't built in by the
  original developer.
\item
  \emph{Gentle slope}: Customization tools should offer a ``gentle
  slope'' from normal use to customization. We explore how our approach
  offers end users opportunities to start customizing software with
  almost no effort, which we hope can make customization a viable part
  of everyday software usage rather than a rare event.
\end{itemize}

@@

\emph{This section still todo. Borrow heavily from Convivial paper, but
extend with newer results}

\begin{itemize}
\tightlist
\item
  Just use airbnb + expedia from Convivial paper?
\item
  instacart + hacker news?
\item
  summarize/mention all 12 sites we've used it with
\item
  real usage:

  \begin{itemize}
  \tightlist
  \item
    1 externally contributed adapter
  \item
    flux adapter in actual use by Sergio
  \end{itemize}
\end{itemize}

@@

In order to promote uniformity within the system, we've defined a small
abstract API, which all table adapters conform to. (Todo: formalize this
more, make it more specific?)

\emph{Output: data table}:The main purpose of a table adapter is to
return a table representing some data source. The table is
live-updating: a table adapter can produce a new version of its data
table at any point (e.g., updating the data table in response to a
change in a website's DOM). When this happens the new data is pushed to
the editor and the query results are refreshed.

@@

\emph{Input: record edits}:The editor can request to a table adapter to
make an edit to a record. The meaning of making an edit can vary
depending on the adapter: in the local storage adapter, an annotation
can be persisted into local storage; in the DOM scraping adapter, an
edit can represent filling in a form field. An adapter can also mark
values as read-only if it wouldn't be meaningful to edit them; for
example, the DOM scraping adapter typically marks page content as
read-only, except for editable form fields.

\emph{Input: query information}:The query engine also sends each table
adapter information about the entire cross-table query view being
created by the user. This functionality is currently mainly used by the
DOM scraping adapter to make changes to the original web page:

@@

Here we describe the three types of table adapters which we have
implemented so far to power the customizations shown above. Then, to
demonstrate the generality of the table adapter paradigm, we describe
three more hypothetical types of table adapters which could extend the
power of the Wildcard system.

\hypertarget{existing-adapters}{%
\subsubsection{Existing Adapters}\label{existing-adapters}}

A \textbf{DOM scraping adapter} extracts data from the DOM of a web
page, and manipulates the DOM to re-order rows, edit form entries, and
inject annotations. A programmer only needs to write a single function
which returns scraped data and pointers to relevant DOM elements; the
Wildcard framework uses this function to implement the rest of the
needed functionality.

(\emph{do we want to call DOM scraping adapter a ``live'' adapter
because its data is also shown elsewhere\ldots?})

@@

we have intentionally designed the table adapter API to be general
enough to encapsulate other types of data and additional functionality
in the future. Here are three concrete examples of such possibilities:

\textbf{Integrated website adapters}: We have taken pains to design a
customization system which is not limited to web scraping as the only
means for integrating with existing sites, and can also accommodate
first party developers adding support for table-driven customization
directly into their own websites. A DOM Scraping table adapter could be
swapped out for such an ``integrated website adapter,'' which implements
the exact same interface as a scraping adapter but by directly accessing
the internal state of the user interface, without needing any scraping
logic.

While we have not yet created a fully operational integrated website
adapter, we think it is possible to create such adapters for existing
frontend web frameworks which result in minimal effort on the part of
the application developers. For example, we have created an early
prototype of a plugin for the Redux state management library, which uses
the Model-View-Update pattern that represents the entire state of a user
interface as a single centralized object. To configure an integrated
website adapter for such an existing application, the user can specify a
function projecting the centralized application state as a table, and
handlers for how data edits to the table should affect the state.

@@

\begin{itemize}
\tightlist
\item
  formulas: do they go here?
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Our editor is closer to visual SQL query engines like Sieuferd or
  Airtable than a freeform spreadsheet.
\item
  cell editors
\item
  Note that there could be other entire table editors
\end{itemize}

@@

\hypertarget{data-oriented-programming}{%
\subsection{Data-oriented programming}\label{data-oriented-programming}}

@@

\hypertarget{black-box-semantic}{%
\subsection{Black-box + semantic}\label{black-box-semantic}}

@@

\textbf{Black-box customization tools} enable customization without
using official extension APIs, enabling a broader range of
customizations on top of more applications. For example, web browser
extensions have demonstrated the utility of customizing websites through
manipulating the DOM, without needing explicit extension APIs to be
built in. However, black-box customization comes with a corresponding
cost: these tools can typically only operate at a low level of
abstraction, e.g.~manipulating user interface elements. This makes it
harder for end users to write scripts, and makes the resulting scripts
more brittle. (todo: support this claim more, provide examples)

@@

(Footnote this paragraph?) This categorization is slightly
oversimplified, and there are existing customization ecosystems that
span both categories. For example, part of the success of browser
extensions stems from the fact that the DOM encourages the use of
standardized semantic UI elements; even if an app developer doesn't
anticipate customization, merely using of semantic and accessible HTML
creates a collateral benefit of easier extension. Another example of a
middle ground is AppleScript, which provides system-wide low-level APIs
for manipulating GUIs as a means of black-box customization, while also
enabling developers to optionally add application-specific APIs to add
support for semantic customization.

@@

\emph{borrow from convivial paper's discussion of in-place}

@@

Find a way to organize all this:

\begin{itemize}
\tightlist
\item
  our own workshop paper. since then\ldots{}

  \begin{itemize}
  \tightlist
  \item
    fundamentally rearchitected around table adapters
  \item
    evaluated on many more websites
  \item
    more fully describing how the system works
  \end{itemize}
\item
  database GUIs:

  \begin{itemize}
  \tightlist
  \item
    Sieuferd \citep{bakke2016}
  \item
    Airtable \citep{2020a}
  \item
    other similar tools?
  \item
    Liu \& Jagadish: rules for a spreadsheet algebra for database
    queries \citep{liu2009}
  \item
    mention DIRECT MANIPULATION
  \end{itemize}
\item
  browser extensions
\item
  instrumental interaction, Scotty \citep{eagan2011}, Webstrates
\item
  customization research tools: Chickenfoot, Coscripter

  \begin{itemize}
  \tightlist
  \item
    Wildcard, like Chickenfoot, wants to hide HTML from users. But we
    show a structured data view, whereas Chickenfoot shows nothing
  \end{itemize}
\item
  visual customization tools:

  \begin{itemize}
  \tightlist
  \item
    Mac Automator
  \item
    Zapier
  \item
    IFTTT
  \item
    These are weakly direct manipulation. But usually the thing you
    directly manipulate in these UIs is the \emph{script}: not the
    actual \emph{data} you want to operate on. VERY DIFFERENT.
  \end{itemize}
\item
  desktop customization: Applescript, VBA, COM
\item
  browser dev tools
\item
  database GUIs
\item
  spreadsheets
\item
  ScrAPIr
\end{itemize}

>>>>

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Customizing Software by Direct Manipulation of Structured Data}

@@

We have known for decades about an alternate approach: \emph{direct
manipulation} \citep{shneiderman1983}, where ``visibility of the object
of interest'' replaces ``complex command language syntax''. Direct
manipulation is the de facto standard in GUIs today, but when it comes
time to customize software, it is rarely to be found. In this work, we
ask: what would it look like to build a software customization interface
that deeply relies on direct manipulation? We take inspiration from
spreadsheets and visual database query interfaces
\citep{2020a, bakke2016}, which enable users to construct queries, edit
data, and compute derived values through direct manipulation. These
tools have successfully enabled millions of end users to compute with
data more easily than ever before.

In this paper we present a technique called \emph{table-driven
customization} which applies these ideas from visual query interfaces in
the context of software customization. We provide the end user with a
table view where they can directly see and manipulate the internal
structured data of an application. The table view and the original UI
serve as coordinated \emph{multiple representations} of the same
underlying state. Changes in the table view result in immediate
corresponding changes to the original user interface of the application,
enabling the user to customize an application with live feedback.

To demonstrate the viability of this approach, we have developed a
browser extension called Wildcard which implements table-driven
customization in the context of the Web. Wildcard works with existing
applications by using web scraping techniques to extract relevant state
into a table, and to reify changes back into the original application
UI. In Section~\ref{sec:examples} we present a many examples of using
Wildcard to implement useful customizations across real websites,
ranging from sorting lists of data to adding whole new features to
applications.

In Section~\ref{sec:architecture} we explain the architecture of the
Wildcard system, focusing on \emph{table adapters} which define the
semantics of how a data table is connected to some underlying state. We
explain how the abstractions of table-driven customization are general
enough to support many customizations beyond the ones we've built so far
in Wildcard. We also explore how applications themselves could be
re-architected to expose tabular data to end users, rather than relying
on web scraping techniques.

@@

\begin{itemize}
\tightlist
\item
  \emph{Customization via direct manipulation}: End users should be able
  to customize an application by examining and modifying its data,
  rather than by issuing imperative commands to an API.
\item
  \emph{Third-party semantic wrappers}: Typically, ``black-box''
  customization tools, which don't rely on official extension APIs,
  resort to offering low-level APIs for customization. Instead, we
  suggest a third-party community of semantic wrappers, offering end
  users a more convenient semantic API for customization, even if it
  wasn't built in by the original developer.
\item
  \emph{Gentle slope}: Customization tools should offer a ``gentle
  slope'' from normal use to customization. We explore how our approach
  offers end users opportunities to start customizing software with
  almost no effort, which we hope can make customization a viable part
  of everyday software usage rather than a rare event.
\end{itemize}

@@

To concretely illustrate how table-driven customization works, here are
several examples of using the Wildcard browser extension, which is
implemented in Typescript and works with Chrome, Firefox, and Edge.

\begin{figure*}
\hypertarget{fig:airbnb-demo}{%
\centering
\includegraphics{media/airbnb-demo-300dpi.png}
\caption{Using Wildcard to augment the Airbnb search page for booking
accommodations}\label{fig:airbnb-demo}
}
\end{figure*}

\hypertarget{augmenting-search-results}{%
\subsection{Augmenting search results}\label{augmenting-search-results}}

In 2012, the travel site Airbnb removed the ability to sort
accommodation searches by price. Users could still filter by price
range, but could no longer view the cheapest listings first. Many users
complained that the change seemed hostile to users. ``It's so
frustrating! What is the logic behind not having this function?'' said
one user on the
\href{https://community.withairbnb.com/t5/Hosting/Sorting-listing-by-price/td-p/559404}{Airbnb
support forum}. Alas, the feature remains missing to this day.

Using Wildcard, the user can fix this omission, while leaving the page's
design and the rest of its functionality unchanged.{
Figure~\ref{fig:airbnb-demo} shows an overview of augmenting the Airbnb
site.} First, the user opens up the Wildcard panel, which shows a table
corresponding to the search results in the page. As they click around in
the table, the corresponding row in the page is highlighted to indicate
the mapping between the views.

Then, the user clicks on the price column header to sort the spreadsheet
and the Airbnb UI by price{ (Figure~\ref{fig:airbnb-demo}, Note A)}.
They also filter to listings with a user rating above 4.5 (another
feature missing in the original Airbnb UI).

After manipulating the data, the user closes the table view and
continues using the website. Because the application's UI usually has a
nicer visual design than a spreadsheet, Wildcard does not aim to replace
it---at any time, the user can use either the UI, the spreadsheet, or
both together.

Many websites that show lists of data also offer actions on rows in the
table, like adding an item to a shopping cart. Wildcard has the ability
to make these ``row actions'' available in the data table through the
site adapter. In the Airbnb UI, saving multiple listings to a Favorites
list requires tediously clicking through them one by one. Using Wildcard
row actions, the user can select multiple rows and favorite all of them
with a single click{ (Figure~\ref{fig:airbnb-demo}, Note B)}. Similarly,
the user can open the detailed pages for many listings at once.

Next, the user wants to jot down some notes about each listing. To do
this, they type some notes into an additional column in each row, and
the notes appear inside the listings in the original UI{
(Figure~\ref{fig:airbnb-demo}, Note C)}. The annotations are saved in
the browser and associated with the backend ID of the listing, so they
will appear in future browser sessions that display the same listing.

Wildcard also includes a formula language that enables more
sophisticated customizations. When traveling without a car, it's useful
to evaluate potential places to stay based on how walkable the
surroundings are. Using a formula, the user can integrate Airbnb with
Walkscore, an API that rates the walkability of any location on a 1-100
scale. When the user calls the \texttt{walkscore} formula with the
listing's latitude and longitude as arguments, it returns the walk score
for that location and shows it as the cell value. Because the cell's
contents are injected into the page, the score also appears in the UI{
(Figure~\ref{fig:airbnb-demo}, Note D)}.

\hypertarget{snoozing-todos}{%
\subsection{Snoozing todos}\label{snoozing-todos}}

\begin{figure*}
\hypertarget{fig:airbnb-demo}{%
\centering
\includegraphics{media/todomvc-demo-300dpi.png}
\caption{Using Wildcard to add a "snooze" feature to the TodoMVC todo list app}\label{fig:todomvc-demo}
}
\end{figure*}

In addition to fetching data from other sources, Wildcard formulas can
also perform computations. In this example, the user would like to
augment the TodoMVC todo list app with a ``snooze'' feature, which will
temporarily hide a todo from the list until a certain date.{
Figure~\ref{fig:todomvc-demo} shows an overview of this customization.}

The user opens the table view, which shows the text and completed status
of each todo. They start the customization by adding a new column to
store the snooze date for each todo{ (Figure~\ref{fig:todomvc-demo},
Note A)}.

The next step is to hide snoozed todos. The user creates a
\texttt{snoozed?} column, which uses a formula to compute whether a todo
is snoozed---i.e., whether it has a snooze date in the future{
(Figure~\ref{fig:todomvc-demo}, Note B)}. Then, they simply filter the
table to hide the snoozed todos{ (Figure~\ref{fig:todomvc-demo}, Note
C)}.

Because the built-in \texttt{NOW()} function always returns the current
datetime, snoozed todos will automatically appear once their snooze date
arrives.

Because this implementation of snoozing was built on the spreadsheet
abstraction, it is completely decoupled from this particular todo list
app. We envision that users could share these types of customizations as
generic browser extensions, which could be applied to any site supported
by Wildcard with no additional effort.

\hypertarget{adding-a-custom-datepicker}{%
\subsection{Adding a custom
datepicker}\label{adding-a-custom-datepicker}}

\begin{figure*}
\hypertarget{fig:airbnb-demo}{%
\centering
\includegraphics{media/expedia-demo-300dpi.png}
\caption{Using Wildcard to augment the Expedia page for booking a flight}\label{fig:expedia-demo}
}
\end{figure*}

It might seem that Wildcard is only useful on websites that display
lists of tabular data, but the table metaphor is flexible enough to
represent many types of data. For example, a flight search form on
Expedia can be represented as a single row, with a column corresponding
to each input{ (Figure~\ref{fig:expedia-demo}, Note A)}.

In some of the previous examples, the table cells were read-only
(because users can't, for example, change the name or price of an Airbnb
listing). In this case, the cells are writable, which means that changes
in the table are reflected in the form inputs. This becomes especially
useful when combined with GUI widgets that can edit the value of a table
cell.

Filling in dates for a flight search often requires a cumbersome
workflow: open up a separate calendar app, find the dates for the trip,
and then carefully copy them into the form. In Wildcard, the user can
avoid this by using a datepicker widget that shows the user's personal
calendar events{ (Figure~\ref{fig:expedia-demo}, Note B)}. The user can
directly click on the correct date, and it gets inserted into both the
spreadsheet and the original form.

@@

(\emph{Todo: formalize this more? Would math notation help?})

In order to promote uniformity within the system, we've defined a small
abstract API, which all table adapters conform to.

The main purpose of a table adapter is to return a table representing
some data source. The table is live-updating: a table adapter can
produce a new version of its data table at any point (e.g., updating the
data table in response to a change in a website's DOM). When this
happens the new data is pushed to the editor and the query results are
refreshed.

@@

The editor can request to a table adapter to make an edit to a record.
The meaning of making an edit can vary depending on the adapter: in the
local storage adapter, an annotation can be persisted into local
storage; in the DOM scraping adapter, an edit can represent filling in a
form field. An adapter can also mark values as read-only if it wouldn't
be meaningful to edit them; for example, the DOM scraping adapter
typically marks page content as read-only, except for editable form
fields.

The query engine also sends each table adapter information about the
entire cross-table query view being created by the user. This
functionality is currently mainly used by the DOM scraping adapter to
make changes to the original web page:

@@

Here we describe in more detail the table adapters which we have
implemented to power the customizations shown above.

\textbf{DOM scraping adapters} are a key part of the Wildcard system.
These adapters extract data from the DOM of a web page, and manipulate
the DOM to re-order rows, edit form entries, and inject annotations. To
create a DOM scraping adapter for a specific website, a programmer does
not need to implement all this functionality from scratch. They only
need to write a single scraping function which uses CSS selectors and
DOM APIs to extract and return relevant elements. The Wildcard framework
then wraps this function to implement the rest of the needed
functionality.

To determine the viability of writing DOM scraping adapters, we built
adapters for 10 websites, including transactional sites like Amazon and
Uber Eats, media consumption sites like Hacker News and Youtube, and a
medical results entry platform managed by a member of our research
group. The adapters ranged from 36 to 117 lines of code, averaging 68
lines. In addition, an external developer unaffiliated with the project
contributed one adapter, designed to sort the Github page listing a
user's repositories, and described the experience as ``very
straightforward.''

Some of the challenges of writing a DOM scraping adapter are the same
ones involved with writing normal web scraping code, but the more
interactive nature of Wildcard introduces additional challenges. One
challenge is triggering updates to the spreadsheet data in response to
UI changes that happen after initial page load. Site adapters are
responsible for recognizing these changes by observing the DOM. So far,
we have been able to use event listeners and the MutationObserver API to
successfully observe changes, but it may prove challenging to observe
changes on some sites only through the DOM. Another challenge is
persisting updates to the DOM---some websites use virtual DOM frameworks
that can occasionally overwrite changes made by Wildcard. In practice,
we've managed to create an implementation that works around these issues
and results in a usable customization experience for all the websites
we've tried.

(\emph{todo: get into the Sorting Problem? i.e., once we sort the page,
the original sort order is lost from the page. Maybe not
necessary\ldots{}})

\hypertarget{other-table-adapters}{%
\subsection{Other table adapters}\label{other-table-adapters}}

@@

We have designed the table adapter API to be general enough to
additional types of functionality in the future. Here are three concrete
examples of hypothetical possibilities:

\textbf{Integrated website adapters}: A key benefit of the table adapter
abstraction is that Wildcard is not coupled to web scraping as the only
means for integrating with existing sites, but can also accommodate
first party developers adding support directly into their own websites.
A DOM Scraping table adapter could be swapped out for such an
``integrated website adapter,'' which implements the exact same
interface as a scraping adapter but by directly accessing the internal
state of the user interface, without needing any scraping logic.

We think it is possible to create such adapters for existing frontend
web frameworks which result in minimal effort on the part of the
application developers. For example, we have created an initial
prototype of a plugin for the Redux state management library. Redux
implements the Model-View-Update pattern, which represents the entire
state of a user interface as a single centralized object. To configure
an integrated website adapter for such an existing application, the user
can simply specify a function projecting the centralized application
state as a table, and define handlers for how data edits to the table
should affect the state.

@@

The query engine is also responsible for executing formulas. We have
built a small language resembling the formula languages in visual query
tools like SIEUFERD. As in those tools, and unlike in spreadsheets,
formulas automatically apply to an entire column of data, and can
reference columns rather than only specific cells.

@@

In Wildcard we provide a basic table editor as the user interface on top
of the query engine. It is built with the Handsontable Javascript
library, which provides UI elements for viewing and editing a table, as
well as basic query operations like sorting and filtering.

\emph{todo: other table instruments}

@@

\hypertarget{customization-by-direct-manipulation}{%
\subsection{Customization by direct
manipulation}\label{customization-by-direct-manipulation}}

@@

Add here:

\begin{itemize}
\tightlist
\item
  multiple representations
\item
  a discussion of
\end{itemize}

\hypertarget{third-party-semantic-wrappers}{%
\subsection{Third-party semantic
wrappers}\label{third-party-semantic-wrappers}}

@@

\textbf{Third-party customization tools} enable customization without
using official extension APIs, enabling a broader range of
customizations on top of more applications. For example, web browser
extensions have demonstrated the utility of customizing websites through
manipulating the DOM, without needing explicit extension APIs to be
built in. However, third-party customization comes with a corresponding
cost: these tools can typically only operate at a low level of
abstraction, e.g.~manipulating user interface elements. This makes it
harder for end users to write scripts, and makes the resulting scripts
more brittle. (todo: support this claim more, provide examples)

@@

(\emph{Footnote this paragraph?}) This categorization is slightly
oversimplified, and there are existing customization ecosystems that
span both categories. For example, part of the success of browser
extensions stems from the fact that the DOM encourages the use of
standardized semantic UI elements; even if an app developer doesn't
anticipate customization, merely using of semantic and accessible HTML
creates a collateral benefit of easier extension. Another example of a
middle ground is AppleScript, which provides system-wide low-level APIs
for manipulating GUIs as a means of black-box customization, while also
enabling developers to optionally add application-specific APIs to add
support for semantic customization.

@@

\emph{borrow from convivial paper's discussion of in-place toolchain}

@@

\begin{itemize}
\tightlist
\item
  our own workshop paper \citep{litt2020}. since then\ldots{}

  \begin{itemize}
  \tightlist
  \item
    fundamentally rearchitected around table adapters
  \item
    evaluated on many more websites
  \item
    more fully describing how the system works
  \end{itemize}
\item
  customization tools

  \begin{itemize}
  \tightlist
  \item
    browser extensions
  \item
    ScrAPIr
  \item
    Thresher
  \item
    Sifter
  \item
    customization research tools: Chickenfoot, Coscripter
  \item
    Wildcard, like Chickenfoot, wants to hide HTML from users. But we
    show a structured data view, whereas Chickenfoot shows nothing
  \item
    visual customization tools:

    \begin{itemize}
    \tightlist
    \item
      Mac Automator
    \item
      Zapier
    \item
      IFTTT
    \item
      These are weakly direct manipulation. But usually the thing you
      directly manipulate in these UIs is the \emph{script}: not the
      actual \emph{data} you want to operate on. VERY DIFFERENT.
    \end{itemize}
  \item
    desktop customization: Applescript, VBA, COM
  \end{itemize}
\item
  database GUIs:

  \begin{itemize}
  \tightlist
  \item
    Sieuferd \citep{bakke2016}
  \item
    Airtable \citep{2020a}
  \item
    other similar tools?
  \item
    Liu \& Jagadish: rules for a spreadsheet algebra for database
    queries \citep{liu2009}
  \item
    Shneiderman paper
  \item
    spreadsheets
  \end{itemize}
\item
  instrumental interaction, Scotty \citep{eagan2011}, Webstrates
\item
  browser dev tools
\item
  Self, Smalltalk
\item
  multiple representations
\item
  Semantic Web
\end{itemize}
